Studying measurement invariance in factor analysis models via stochastic processes

### 2011-02-04 MIpaper/par_instable.tex
Revised example section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@34 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\fixme{I have explicitly written down all three test statistics: DM because it is
intuitive, max LM because we need it to connect to the previous section, and CvM
because it nicely complements the picture. For the simulations, I think we should
also consider all three tests, maybe also the max LR test. Of course, we shouldn't
make things too complex but we can then still decide which results to choose.}

@@

As an example of the ideas discussed above, consider a
hypothetical battery of six scales administered to students aged 14 to 17
years.  Three of the scales are intended to measure verbal ability, and
three of the scales are intended to measure mathematical ability.  We
may observe a maturation effect in the resulting data, whereby the scales
measure older students better than younger students (i.e., the scales
are too difficult for younger students).  The researcher's goal is to
study whether the scales are measurement invariant with
respect to age, which is taken to be the auxiliary variable $V$.

@@

To formally represent these ideas, we specify that the data arise from
a factor analysis model with two factors.  The base model, displayed in
Figure~\ref{fig:famod}, specifies that measurement invariance holds,
with three scales arising from the verbal factor and three 
scales arising from the mathematical factor.
For measurement invariance violations, we specify that $V$
(student age) impacts the values of 
verbal factor loadings in the model: 
if students are 16 or 17 years of age, then the factor
loadings corresponding to the first 
factor ($\lambda_{11}, \lambda_{21}, \lambda_{31}$) reflect those in
Figure~\ref{fig:famod}.  If students are 14 or 15 years of age,
however, then the factor loadings corresponding to the first factor
are three standard errors lower than those in Figure~\ref{fig:famod}
(the standard errors were obtained by fitting
  the model to a sample of size 200, where the sample consisted of all
  ``older'' subjects).
This violation states that the verbal ability scales
lack measurement invariance with respect to age.  For simplicity, we
assume that the mathematical scales are invariant.

A sample of size 200 was generated from the model described above, and
a test was conducted to examine measurement invariance of the three
verbal scales.  To carry out
the test, a confirmatory factor analysis model (with the paths
displayed in Figure~\ref{fig:famod}) was fit to the
data.  Casewise derivatives and 
the observed information matrix were then obtained, and they were used
to calculate the cumulative score process via
\eqref{eq:cumscore}.  Finally, we
obtained a 
double-max statistic via Equation~\eqref{eq:dmax} and associated
$p$-value via Equation~\eqref{eq:absmax_p}.  Note that, because we
were examining only the three verbal factor loading parameters, the
specific null hypothesis being tested is:
\begin{equation}
    \label{eq:exh0}
H_0:\ (\lambda_{i,11}\ \lambda_{i,21}\ \lambda_{i,31}) =
(\lambda_{11}\ \lambda_{21}\ \lambda_{31}),\ i=1,\ldots,n,
\end{equation}
where $(\lambda_{i,11}\ \lambda_{i,21}\ \lambda_{i,31})$ represent the
verbal factor loading parameters for student $i$.

\subsection{Results}
In the results section, we first describe overall results and then 
describe estimation of $\eta$ and 
isolation of model parameters violating
measurement invariance.\footnote{R code for obtaining these results is
  available at \url{http://semtools.r-forge.r-project.org}.}

\subsubsection{Overall Results}
Results related to the cumulative score process are displayed in
Figure~\ref{fig:cusumex}.  In this figure, each panel represents the
process for a single factor loading, with student age on the x-axis and
cumulative scores on the y-axis.  The dashed horizontal
lines represent the critical value for rejecting $H_0$ at $\alpha=.05$
based on the double-max statistic; if any of the three processes
crosses the horizontal line, then we reject $H_0$.  It is seen that
the processes for both the second and third factor loadings
($\lambda_{21}$ and $\lambda_{31}$) cross the dashed line, indeed
implying rejection of $H_0$.  The absolute maximum across all three processes,
corresponding to the double-max test statistic, is 1.77.  This
statistic is obtained from the minimum of the third
panel (whose absolute value is the maximum), close to an
age of 16.  It achieves a $p$-value of .012.

\subsubsection{Estimation of $\eta$ and Parameter Isolation}
The results of the test imply that the 
verbal scales lack measurement invariance, and the age at which the
maximum is achieved provides the estimate $\hat{\eta}$.
The student corresponding to the maximum
absolute cumulative sum (the solid circle in the third graph) is 15.7
years of age, implying one subgroup of 
students below 15.7 years and one above 15.7 years.  This agrees well
with the true $\eta$ of 16.0 years.

The results of the test also yield information about specific
parameters violating measurement invariance.  We simultaneously tested
three factor loading parameters for measurement invariance.  Given
rejection of the hypothesis of measurement invariance, we can 
do posthoc 
tests to study specific parameters violating measurement invariance.
This involves individual examination of each cumulative score process
(i.e., of each of the three graphs in Figure~\ref{fig:cusumex}),
along with a procedure that controls for Type I error.  Taking a
Bonferroni approach, we can examine whether each process's maximum
surpasses the critical value associated with an $\alpha$ of
$.05/3=.017$.  This amounts to the same critical value as that used in
the overall test, 
meaning that we simply examine the individual processes that cross
the horizontal dashed lines in Figure~\ref{fig:cusumex}.  The figure
shows that the second and third parameters (i.e., $\lambda_{21}$ and
$\lambda_{31}$) cross the dashed line, so we would conclude a measurement
invariance violation with respect to age in the second and third
verbal tests.  The fact that the cumulative score process for
$\lambda_{11}$ did not achieve the critical value represents a Type II
error, which implicitly brings into question the posthoc tests' power.  We
generally address the issue of power in the simulations below.

% NEED TO MODIFY
\begin{figure}
\caption{Cumulative score processes for the verbal factor loadings,
  based on the example involving measurement invariance with respect
  to student age.  Each panel represents the cumulative score process
  for a single factor loading; the red line represents the critical values for 
  the absolute maximum at $\alpha=.05$.}
\label{fig:cusumex}
\includegraphics[height=5in]{example.pdf}
\end{figure}

>>>>

@@

As an example of the ideas discussed above, consider a
hypothetical battery of six scales administered to students aged 13 to 18
years.  Three of the scales are intended to measure verbal ability, and
three of the scales are intended to measure mathematical ability.  We
may observe a maturation effect in the resulting data, whereby the scales
measure older students better than younger students (i.e., younger
students have not yet learned all the material).  The researcher's goal is to
study whether the scales are measurement invariant with
respect to age, which is taken to be the auxiliary variable $V$.

@@

To formally represent these ideas, we specify that the data arise from
a factor analysis model with two factors.  The base model, displayed in
Figure~\ref{fig:famod}, specifies that measurement invariance holds,
with three scales arising from the verbal factor and three 
scales arising from the mathematical factor.
For measurement invariance violations, we specify that $V$
(student age) impacts the values of 
verbal factor loadings in the model: 
if students are 16 through 18 years of age, then the factor
loadings corresponding to the first 
factor ($\lambda_{11}, \lambda_{21}, \lambda_{31}$) reflect those in
Figure~\ref{fig:famod}.  If students are 13 through 15 years of age,
however, then the factor loadings corresponding to the first factor
are three standard errors lower than those in Figure~\ref{fig:famod}
(the standard errors were obtained by fitting
  the model to a sample of size 200, where the sample consisted of all
  ``older'' subjects).
This violation states that the verbal ability scales
lack measurement invariance with respect to age.  For simplicity, we
assume that the mathematical scales are invariant.

A sample of size 200 was generated from the model described above, and
a test was conducted to examine measurement invariance of the three
verbal scales.  To carry out
the test, a confirmatory factor analysis model (with the paths
displayed in Figure~\ref{fig:famod}) was fit to the
data.  Casewise derivatives and 
the observed information matrix were then obtained, and they were used
to calculate the cumulative score process via
\eqref{eq:cumscore}.  Finally, we
obtained various test statistics and $p$-values from the cumulative score
process.  These include the double-max statistic from \eqref{eq:dmax},
the Cram\'{e}r-von Mises statistic from \eqref{eq:cvm}, and the max-LM
statistic from \eqref{eq:maxlm}.

Note that the tests give us the flexibility to test various subsets of
parameters.  For example, were we only interested in the 
three verbal factor loading parameters, we can test
\begin{equation}
    \label{eq:exh0}
H_0:\ (\lambda_{i,11}\ \lambda_{i,21}\ \lambda_{i,31}) =
(\lambda_{11}\ \lambda_{21}\ \lambda_{31}),\ i=1,\ldots,n,
\end{equation}
where $(\lambda_{i,11}\ \lambda_{i,21}\ \lambda_{i,31})$ represent the
verbal factor loading parameters for student $i$.  Alternatively, we
can consider all model parameters (including means), leading to a test
of \eqref{eq:h0}.  We consider both of these tests below.

\subsection{Results}
In the results section, we first describe overall results.  We then 
describe estimation of $\nu$, and 
isolation of model parameters violating
measurement invariance.  R code for obtaining these results is
  available at \url{http://semtools.r-forge.r-project.org}.

\subsubsection{Overall Results}
Test statistics for the hypotheses \eqref{eq:exh0} and~\eqref{eq:h0}
are displayed in  
Figure~\ref{fig:cusumex}.
Each panel displays a test 
statistic's fluctuation across values of student age, with the first
column containing tests of \eqref{eq:exh0} and the second column
containing tests of \eqref{eq:h0}.  The solid
horizontal 
lines represent critical values for $\alpha=.05$, and
the Cram\'{e}r-von Mises panels 
also contain a dashed line depicting the value of the test
statistic (test statistics for the others are simply 
the maxima of the processes).
In other words, for panels in the first and third rows, \eqref{eq:exh0} is
rejected if the process crosses the horizontal line.  For panels in
the second row, \eqref{eq:h0} is rejected if the dashed horizontal line is
higher than the solid horizontal line.

The figures display many test attributes, all of which generally
conform with the underlying theory.  First, when we test
only the three verbal factor loadings, all three tests can detect the
change.  Conversely, when all nineteen model parameters are tested,
only the double-max test detects a change.  This implies that 
power increases as larger proportions of tested parameters
change.  The results also imply that the double-max test is more
sensitive to fluctuations among a small subset of parameters.

Figure~\ref{fig:lrlm} 
compares the max-LM statistic (black line) to the max-LR
statistic (blue line) from \eqref{eq:maxlr}, as applied to testing
\eqref{eq:h0}.  The critical values for these two
tests are identical, hence the single horizontal line.
The figure
shows that the two statistics are very similar to one another, with
both maxima at the dotted vertical line.  This
is generally to be expected, because the two tests are asymptotically
equivalent.  For this specific situation, however, the max-LR
statistic rejects \eqref{eq:h0} while the max-LM statistic does not.
More generally, the 
max-LR statistic cannot be obtained from the empirical fluctuation
process, so it must be computed ``by hand'' (i.e., by computing
the LR statistic over all values of the threshold $\nu$, re-fitting the
factor analysis model for each value).

\subsubsection{Estimation of $\nu$ and Parameter Isolation}
As described above, the tests of \eqref{eq:exh0} imply that the 
verbal scales lack measurement invariance.  
We can also use the tests to: (1) estimate the 
threshold $\nu$, and (2) isolate specific parameters that violate
measurement invariance.  
For example, as described previously, estimates of $\nu$ can be obtained by
examining the peaks in Figure~\ref{fig:cusumex}.  For all six panels
in the figure, the peaks occur near an age of 16.1.  This agrees well
with the true threshold of 16.0.

The double-max test yields specific information about 
parameters violating measurement invariance.  Given
rejection of the hypothesis of measurement invariance \eqref{eq:exh0},
we can do posthoc 
tests to study specific parameters violating measurement invariance.
This involves individual examination of each parameter's cumulative
score process,
along with a procedure that controls for Type I error.  
Figure~\ref{fig:posthoc} shows the individual cumulative score processes for the
verbal factor loadings, with the horizontal lines reflecting the
critical value at $\alpha=.05$.
  The figure
shows that the third parameter (i.e., $\lambda_{31}$) crosses the
dashed line, so we would conclude a measurement 
invariance violation with respect to age for the third
verbal test.  The fact that the cumulative score process for the first
and second loadings did not achieve the critical value represents a Type II
error, which implicitly brings into question the posthoc tests' power.  We
generally address the issue of power in the simulations below.

\begin{figure}
\caption{Three test statistics of \eqref{eq:exh0} and \eqref{eq:h0}, 
  based on the example involving measurement invariance with respect
  to student age.  Solid, horizontal lines represent critical values
  at $\alpha=.05$, and the dotted, horizontal lines (second row)
  represent values of the Cram\'{e}r-von Mises test statistic.}
\label{fig:cusumex}
\includegraphics[height=9in]{gefp_3_19.pdf}
\end{figure}

\begin{figure}
\caption{A comparison of the max-LM (black line) and max-LR (blue line) test
  statistics for \eqref{eq:h0}.  The solid, horizontal line
  corresponds to the critical value at $\alpha=.05$.}
\label{fig:lrlm}
\includegraphics[height=5in]{lmlr_19.pdf}
\end{figure}

\begin{figure}
\caption{Posthoc tests of individual verbal factor loadings, based on
  the absolute maximum of each cumulative score process.  The top
  panel depicts the process for $\lambda_{11}$, the middle panel
  depicts the process for $\lambda_{21}$, and the bottom panel depicts
  the process for $\lambda_{31}$.  The solid, horizontal lines
  correspond to the critical value at $\alpha=.05$.}
\label{fig:posthoc}
\includegraphics[height=5in]{gefp_posthoc.pdf}
\end{figure}
